---
title: "Analysis of Factors Influencing Individual Income Levels"
author: "Group Number: 30"
number-sections: true
format: 
  html:
    embed-resources: true
    code-tools: true
  pdf: default
editor_options: 
  chunk_output_type: console
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

```{r}
#| label: libraries
library(tidyverse)
library(ggplot2)
library(caret)
library(pROC)
library(GGally)
```

# Introduction {#sec-Intro}

This analysis aims to identify key socioeconomic factors that influence whether an individual earns more than \$50,000 per year, using data from the **1994 US Census**. The dataset includes demographic and employment-related variables such as **age, education level, marital status, occupation, sex, hours worked per week, and nationality**, with income categorized into two groups: **≤\$50K and \>\$50K per year**.

To address this, a **Generalized Linear Model (GLM)** will be applied to evaluate the impact of these factors on income levels. The findings will provide insights into the most significant predictors of higher earnings, contributing to a deeper understanding of income distribution patterns. The results will be summarized and presented in a structured format.

# Exploratory Data Analysis {#sec-EDA}

```{r}
#| label: data processing
df <- read.csv("C:\\Users\\Lenovo\\Desktop\\data analysis skills\\dataset30.csv", stringsAsFactors = FALSE)
df <- df %>%
  mutate(across(everything(), ~str_remove_all(., ","))) %>%  
  mutate(Hours_PW = as.numeric(Hours_PW),  
         Age = as.numeric(Age),  
         Income = ifelse(Income == ">50K", 1, 0))  
df <- df %>% filter(Occupation != "?" & Nationality != "?")
df <- df %>% select(Age, Education, Marital_Status, Occupation, Sex, Hours_PW, Income)
df <- df %>% mutate(across(where(is.character), as.factor))
str(df)
df$Age <- scale(df$Age)
set.seed(1111)
trainIndex <- createDataPartition(df$Income, p = 0.8, list = FALSE)
train_data <- df[trainIndex, ]
test_data <- df[-trainIndex, ]
```

This data analysis first involved cleaning, transforming, selecting, and standardizing the raw data. Then, the dataset was split into **80% training set / 20% test set** to ensure data quality and enhance the model's generalization ability.

```{r}
# Age vs. Income
ggplot(df, aes(x = as.factor(Income), y = Age, fill = as.factor(Income))) + 
  geom_boxplot() +
  scale_x_discrete(labels = c("<=50K", ">50K")) +
  ggtitle("Age vs. Income") +
  xlab("Income Category") +
  ylab("Age") +
  theme_minimal()
```

As can be seen from the figure, there is a certain positive correlation between age and income: the median age of the high-income group (income \>50K) is higher than that of the low-income group (income ≤50K), indicating that older people are more likely to has the higher income.

```{r}
#Working Hours vs. Income
ggplot(df, aes(x = as.factor(Income), y = Hours_PW, fill = as.factor(Income))) + 
  geom_boxplot() +
  scale_x_discrete(labels = c("<=50K", ">50K")) +
  ggtitle("Working Hours vs. Income") +
  xlab("Income Category") +
  ylab("Hours Per Week") +
  theme_minimal()
```

Working hours have a significant impact on income, higher earners tend to work longer hours, but hours alone do not fully determine income. The working hours of low-income groups are more concentrated, but there are more outliers: individuals with extremely low working hours (\<25 hours) and extremely high working hours (\>50 hours).

```{r}
#Education Level vs. Income (Bar Chart)
ggplot(df, aes(x = Education, fill = as.factor(Income))) + 
  geom_bar(position = "fill") +  
  coord_flip() +  
  scale_fill_manual(values = c("skyblue", "orange"), labels = c("<=50K", ">50K")) +
  ggtitle("Education Level vs. Income") +
  xlab("Education Level") +
  ylab("Proportion") +
  theme_minimal()
```

Education level is positively correlated to income, and higher degree is more likely to earn a high income. In particular, the proportion of high income of master's, doctor's and professional school graduates is significantly higher, while those with low education level are mainly concentrated in low-income groups. This suggests that education level plays an important role in income.

```{r}
#Occupation vs. Income (Bar Chart)
ggplot(df, aes(x = Occupation, fill = as.factor(Income))) + 
  geom_bar(position = "fill") + 
  coord_flip() +
  scale_fill_manual(values = c("lightgreen", "red"), labels = c("<=50K", ">50K")) +
  ggtitle("Occupation vs. Income") +
  xlab("Occupation") +
  ylab("Proportion") +
  theme_minimal()
```

The figure shows clear income differences between different occupational categories. A higher percentage of "Exec-managerial" and "Prof-specialty" occupations make more than 50,000, indicating that these occupations are more likely to generate higher income. By contrast, the proportion of "Priv-house-serv" and "Handlers-cleaners" is significantly lower.

```{r}
#Gender vs. Income (Bar Chart)
ggplot(df, aes(x = Sex, fill = as.factor(Income))) + 
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("purple", "pink"), labels = c("<=50K", ">50K")) +
  ggtitle("Gender vs. Income") +
  xlab("Gender") +
  ylab("Proportion") +
  theme_minimal()
```

There are significant differences in income distribution between men and women. The propotion of men earning more than 50,000 is significantly higher than that of women, while the women income less than 50,000 is relatively higher.

# Formal Data Analysis {#sec-FDA}

```{r}
#| label: Model Training
full_model <- glm(Income ~ ., data = train_data, family = binomial)
stepwise_model <- step(full_model, direction = "backward")
```

The **logistic regression** model is built to predict income level **(`Income`)** by considering various socioeconomic factors, such as **age, education, marital status, occupation, sex,** and **hours worked per week**. To further refine the model, **stepwise regression** is applied using **backward elimination**, which systematically removes variables that do **not significantly** **(sex)** contribute to the prediction based on the **Akaike Information Criterion (AIC)**. The initial model's AIC was **883.77**, while the optimized model's AIC was reduced to **882.19**, indicating that the model became more concise while maintaining a good fit. As a result, this approach not only simplifies the model but also helps to reduce overfitting and enhance interpretability by retaining only the most significant predictors.

```{r}
#| label: Performance Evaluation
test_predictions <- predict(stepwise_model, newdata = test_data, type = "response")
test_predicted_classes <- ifelse(test_predictions > 0.5, 1, 0)
confusionMatrix(factor(test_predicted_classes), factor(test_data$Income))
conf_matrix <- confusionMatrix(factor(test_predicted_classes), factor(test_data$Income))
accuracy <- conf_matrix$overall["Accuracy"]
print(accuracy)
roc_curve_test <- roc(test_data$Income, test_predictions)
plot(roc_curve_test, col = "red", main = "ROC Curve for Test Data", xlim = c(1, 0), ylim = c(0, 1), asp = 1)
auc(roc_curve_test)
```

The evaluation results of this logistic regression model on the test set indicate a high classification capability, with an accuracy of **85.41%** within the **95% confidence interval (80.73% - 89.32%)**, demonstrating stable predictive performance. The confusion matrix further shows that the model has a strong ability to identify low-income individuals, achieving a **sensitivity of 94.86%**, meaning that most low-income individuals are correctly classified. However, the **specificity is only 55.22%**, suggesting that the model has some difficulty in correctly identifying high-income individuals, with a considerable number being misclassified as low-income. Despite this, the model's **accuracy (ACC)** still reaches **85.4%**. In terms of overall classification performance, the **AUC value of 0.8683** indicates that the model performs well in distinguishing between high-income and low-income individuals, and the **ROC curve** demonstrates strong discriminatory power.

# Conclusions {#sec-Conc}

This study utilized data from the **1994 U.S. Census** to analyze key socioeconomic factors influencing individual income levels. The results indicate that **education level, occupation, marital status, and age** are crucial determinants of income, with higher education and specialized professions significantly increasing earning potential. Additionally, **sex**, as men are more likely to earn higher incomes. Furthermore, **weekly working hours** have a certain impact on income, though with diminishing marginal returns.
